---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 22 @ 11:59PM
author: "Zehan Hu"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 

1. Data preprocessing and feature engineering.


```{r}

library(tidyverse)
library(tidymodels)
mimiciv_icu_cohort<-readRDS("../hw4/mimiciv_shiny/mimic_icu_cohort.rds")

```





```{r}
mimiciv_icu_cohort <- mimiciv_icu_cohort |>
  select(subject_id, hadm_id, stay_id,"Non Invasive Blood Pressure diastolic"="as.numeric(\"Non Invasive Blood Pressure diastolic\")",
         "Non Invasive Blood Pressure systolic"="as.numeric(\"Non Invasive Blood Pressure systolic\")",
         "Heart Rate"="as.numeric(`Heart Rate`)",temperature_fahrenheit="as.numeric(`Temperature Fahrenheit`)",los_long, insurance,language,marital_status,race,hospital_expire_flag,gender,sodium,chloride,creatinine,potassium,glucose,hematocrit,wbc,bicarbonate)
```

```{r}
# Assuming your data frame is named df

# Convert all character columns to factors
mimiciv_icu_cohort[] <- lapply(mimiciv_icu_cohort, function(x) if(is.character(x)) factor(x) else x)

mimiciv_icu_cohort$los_long<-as.factor(mimiciv_icu_cohort$los_long)

```




2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.
```{r}
#| eval: false
set.seed(203)

# sort
mimiciv_icu_cohort <- mimiciv_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id)

data_split <- initial_split(
  mimiciv_icu_cohort, 
  # stratify by los_long
  strata = "los_long", 
  prop = 0.5
  )
```


```{r}
mimic_other <- training(data_split)
dim(mimic_other)
```

```{r}
mimic_test<-testing(data_split)
dim(mimic_test)

```


```{r}
sapply(mimic_other, class)
```


```{r}

norm_recipe <- 
  recipe(
    los_long ~ .,
    data = mimic_other
  ) %>%
  step_impute_mean(all_numeric(), -all_outcomes()) %>%
  # Impute categorical variables using KNN or another appropriate method
  step_impute_knn(all_nominal(), -all_outcomes()) %>%
  # Convert all nominal variables to dummy variables, excluding the outcome
  step_dummy(all_nominal(), -all_outcomes()) %>%
  # Remove zero variance predictors
  step_zv(all_numeric_predictors()) %>%
  # Normalize all numeric predictors
  step_normalize(all_numeric_predictors())


norm_recipe
```




### Random Forest



```{r}
rf_mod <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) %>% 
  set_engine("ranger")
rf_mod
rf_wf <- workflow() %>%
  add_recipe(norm_recipe) %>%
  add_model(rf_mod)
rf_wf

param_grid <- grid_regular(
  trees(range = c(100L, 300L)), 
  mtry(range = c(1L, 5L)),
  levels = c(3, 5)
  )
param_grid
```






```{r}
set.seed(203)

folds <- vfold_cv(mimic_other, v = 5)
folds
```


```{r}
rf_fit <- rf_wf %>%
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
rf_fit
```


```{r}
rf_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  mutate(mtry = as.factor(mtry)) %>%
  ggplot(mapping = aes(x = trees, y = mean, color = mtry)) +
  # geom_point() + 
  geom_line() + 
  labs(x = "Num. of Trees", y = "CV AUC")
```

```{r}
rf_fit %>%
  show_best("roc_auc")
best_rf <- rf_fit %>%
  select_best("roc_auc")
best_rf
```


```{r}
# Final workflow
final_wf <- rf_wf %>%
  finalize_workflow(best_rf)
final_wf
# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf %>%
  last_fit(data_split)
final_fit
```

```{r}
# Test metrics
final_fit %>% 
  collect_metrics()
```


### XG boosting


```{r}
gb_mod <- 
  boost_tree(
    mode = "classification",
    trees = 1000, 
    tree_depth = tune(),
    learn_rate = tune()
  ) %>% 
  set_engine("xgboost")
gb_mod
```


```{r}
gb_wf <- workflow() %>%
  add_recipe(norm_recipe) %>%
  add_model(gb_mod)
gb_wf
```


```{r}
param_grid <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 2), trans = log10_trans()),
  levels = c(3, 10)
  )
param_grid
```


```{r}
set.seed(203)

folds <- vfold_cv(mimic_other, v = 5)
folds
```

```{r}
gb_fit <- gb_wf %>%
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
gb_fit
```


```{r}
gb_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  mutate(tree_depth = as.factor(tree_depth)) %>%
  ggplot(mapping = aes(x = learn_rate, y = mean, color = tree_depth)) +
  geom_point() +
  geom_line() +
  labs(x = "Learning Rate", y = "CV AUC") +
  scale_x_log10()
```


```{r}
gb_fit %>%
  show_best("roc_auc")
best_gb <- gb_fit %>%
  select_best("roc_auc")
best_gb
```


```{r}
# Final workflow
final_wf <- gb_wf %>%
  finalize_workflow(best_gb)
final_wf
# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf %>%
  last_fit(data_split)
final_fit
```


```{r}
# Test metrics
final_fit %>% 
  collect_metrics()
```



### SVM


```{r}

```




3. Train and tune the models using the training set.

4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?
